---
title: "LineagePulse"
author: "Boying"
date: "11/30/2017"
output: html_document
---

```{r setup, include=FALSE}
# run from command line with : Rscript -e "rmarkdown::render('LineagePulse.Rmd')"
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message=FALSE}
library(LineagePulse)
library(gridExtra)
library(NMF)
library(VennDiagram)
library(ggplot2)
```

```{r Functions}
# to be used for the colors of the heatmap:
seqPal2 <- colorRampPalette(c("orange", "black", "blue"))(16)
seqPal2 <- (c("yellow", "gold2", seqPal2))
seqPal2 <- rev(seqPal2)

plotGenes <- function(objLP, n, type = "mostSig"){
  if (type == "mostSig"){
    orderFac <- 1
  } else if (type == "leastSig"){
    orderFac <- -1
  }
  ordered_results <- objLP$dfResults
  ordered_results <- ordered_results[order(orderFac*ordered_results$padj), ]

  mostSigGenes <- list()
  for (i in 1:n){
    mostSigGenes[[i]] <- plotGene(objLP = objLP, boolLogPlot = FALSE,
                               strGeneID=ordered_results[i,]$gene,
                               boolLineageContour = FALSE) + theme_bw()
  }
  do.call("grid.arrange", c(mostSigGenes, ncol=3))
}

propSig <- function(objLP){
  qvals <- objLP$dfResults$padj
  thresholds <- seq(-240, -0, by = 2)
  props <- sapply(10^thresholds, function(x) mean(qvals <= x))
  cat(paste0(round(props[length(props)-1]*100, 2), "% of the genes are called significant when we require the adjusted p-value to be smaller than 0.01"))
  
  plot(thresholds, props, cex = 0.3, 
       xlab = "log10(q-value)", ylab = "Proportion", 
       main = "Proportion of genes called significant")
  lines(thresholds, props)
}

sigGenesHeatmap <- function(objLP, lineageIdx, scale = TRUE, nGenes, method){
  res <- objLP$dfResults
  res <- res[order(res$padj), ]
  sig_genes <- res[1:nGenes, "gene"]
  cells <- psdt[!is.na(psdt[, paste0("curve", lineageIdx)]), ]
  cells <- cells[order(cells[, paste0("curve", lineageIdx)]), ]
  rawCountsSubset <- rawCounts[sig_genes, rownames(cells)]
  logRawCountsSubset <- log(rawCountsSubset+1)
  if (scale){
    logRawCountsSubset <- t(scale(t(logRawCountsSubset),scale=TRUE,center=FALSE))
    # t(apply(logRawCountsSubset, 1, function(x) x/max(x)))
    m <- max(logRawCountsSubset)
    qnt <- quantile(logRawCountsSubset, 0.99)
    brk <- seq(0, qnt, length = 50) 
    brk <- c(brk, m) 
    aheatmap(logRawCountsSubset, Colv = NA, breaks = brk, col = seqPal2, Rowv = FALSE, labRow = NA, labCol = NA, 
             annCol = data.frame(time = cells[, paste0("curve", lineageIdx)]),
             main = method, legend=FALSE, annLegend = FALSE)
  } else {
    aheatmap(logRawCountsSubset, Colv = NA, col = seqPal2, Rowv = FALSE, labRow = NA, labCol = NA, 
             annCol = data.frame(time = cells[, paste0("curve", lineageIdx)]),
             main = method, legend=FALSE, annLegend = FALSE)
  }
  return(brk)
}
```

```{r loadData}
normalizedCounts <- as.matrix(read.table("data/normalisedCountsVariableGenes.txt", header=TRUE, sep = ""))
rawCounts <- read.table("data/rawCounts.txt", header=TRUE, sep = "")
rownames(rawCounts) <- rawCounts$ID
rawCounts$ID <- NULL
rawCounts <- as.matrix(rawCounts)
# subset the raw counts
rawCounts <- rawCounts[rownames(normalizedCounts), colnames(normalizedCounts)]
# load the pseudo time data
load("output/pseudotimes.RData")
rownames(psdt) <- colnames(normalizedCounts)
```

```{r}
# Load data
load("output/PvalList.RData")
PvalP.adj <- as.data.frame(PvalP.adj)
PvalP.adj$gene <- rownames(normalizedCounts)
PvalNP.adj <- as.data.frame(PvalNP.adj)
PvalNP.adj$gene <- rownames(normalizedCounts)
```

```{r}
sigGenesHeatmapLoess <- function(Pval.adj, lineageIdx, scale = TRUE, nGenes, method, brk, drawLegend){
  Pval.adj <- Pval.adj[order(Pval.adj[, lineageIdx]), ]
  sig_genes <- Pval.adj[1:nGenes, "gene"]
  cells <- psdt[!is.na(psdt[, paste0("curve", lineageIdx)]), ]
  cells <- cells[order(cells[, paste0("curve", lineageIdx)]), ]
  rawCountsSubset <- rawCounts[sig_genes, rownames(cells)]
  logRawCountsSubset <- log(rawCountsSubset+1)
  if (scale){
    logRawCountsSubset <- t(scale(t(logRawCountsSubset),scale=TRUE,center=FALSE))
    aheatmap(logRawCountsSubset, Colv = NA, breaks = brk, col = seqPal2,  Rowv = FALSE, labRow = NA, labCol = NA, 
             annCol = data.frame(time = cells[, paste0("curve", lineageIdx)]),
             main = method, legend=drawLegend, annLegend = drawLegend)
  } else {
    aheatmap(logRawCountsSubset, Colv = NA, col = seqPal2,  Rowv = FALSE, labRow = NA, labCol = NA, 
             annCol = data.frame(time = cells[, paste0("curve", lineageIdx)]),
             main = method, legend=drawLegend, annLegend = drawLegend)
  }
}
```


# Lineage 1


```{r LoadDataL1}
load("output/lineage1.RData")
results <- as.data.frame(objLP$dfResults)
results <- data.frame(gene = results$gene,
                      pval_adjusted = results$padj)
save(results, file = "output/lineage1_results.RData")
# head(objLP$dfResults)
```

### Percentage of genes called significant as we change q-value threshold

```{r fig.width=5, fig.height=5, out.width='50%'}
propSig(objLP)
```

### Heatmap of significant genes

Scaled log expression of lineage 1. Heatmap of top 500 significant genes detected by three methods.

```{r, message=FALSE, fig.width=10, fig.height=7, out.width='100%'}
par(mfrow=c(1, 3))
layout(matrix(c(1,2,3), 1), widths = c(2, 2.7))
brk = sigGenesHeatmap(objLP, 1, TRUE, 500, "LineagePulse")
sigGenesHeatmapLoess(PvalP.adj, 1, TRUE, 500, "Loess", brk, TRUE)
# sigGenesHeatmapLoess(PvalNP.adj, 1, TRUE, 500, "Loess Nonparametric", brk, TRUE)
```

Log expression of lineage 1. Heatmap of top 500 significant genes detected by three methods.

```{r, message=FALSE, fig.width=10, fig.height=7, out.width='100%'}
par(mfrow=c(1, 3))
layout(matrix(c(1,2,3), 1), widths = c(2, 2.7))
brk = sigGenesHeatmap(objLP, 1, FALSE, 500, "LineagePulse")
sigGenesHeatmapLoess(PvalP.adj, 1, FALSE, 500, "Loess", brk, TRUE)
# sigGenesHeatmapLoess(PvalNP.adj, 1, FALSE, 500, "Loess Nonparametric", brk, TRUE)
```

### Visualization of 9 most significant genes

```{r fig.width=20, fig.height=12, out.width='100%'}
plotGenes(objLP, 9, "mostSig")
```

### Visualization of 9 least significant genes

```{r fig.width=20, fig.height=12, out.width='100%', warning=FALSE}
plotGenes(objLP, 9, "leastSig")
```

### Venn Diagram

```{r}
nGenes = 500
lp = objLP$dfResults
lp = lp[order(lp$padj), ]
lp = lp[1:nGenes, "gene"]
loessP = PvalP.adj[order(PvalP.adj$L1), ]
loessP = loessP[1:nGenes, "gene"]
loessNP = PvalNP.adj[order(PvalNP.adj$L1), ]
loessNP = loessNP[1:nGenes, "gene"]

  tmp <- venn.diagram(
    x = list(
      "LineagePulse" = lp, 
      "Loess" = loessP
    ),
    filename = paste0("Lineage1.png"),
        output = TRUE ,
        imagetype="png" ,
        compression = "lzw",
   col = "black",
    lty = "dotted",
    lwd = 2.5,
        height = 700 , 
        width = 700 , 
        resolution = 300,
        fill = c("cornflowerblue", "green"),    
    alpha = 0.40,
    cat.cex = 0.6,
   cex = 1.5,
    main.cex = 1.5,
    sub.cex = 1.25,
    cat.fontfamily = "serif"
  )
```


```{r}
nGenes = 500
lp = objLP$dfResults
lp = lp[lp$padj <= 0.01, "gene"]
loessP = PvalP.adj[PvalP.adj$L1 <= 0.01, "gene"]

  tmp <- venn.diagram(
    x = list(
      "LineagePulse" = lp, 
      "Loess" = loessP
    ),
    filename = paste0("Lineage1Sig.png"),
        output = TRUE ,
        imagetype="png" ,
        compression = "lzw",
   col = "black",
    lty = "dotted",
    lwd = 2.5,
        height = 700 , 
        width = 700 , 
        resolution = 300,
        fill = c("cornflowerblue", "green"),    
    alpha = 0.40,
    cat.cex = 0.6,
   cex = 1.5,
    main.cex = 1.5,
    sub.cex = 1.25,
    cat.fontfamily = "serif"
  )
```

# Lineage 2


```{r LoadDataL2}
load("output/lineage2.RData")
results <- as.data.frame(objLP$dfResults)
results <- data.frame(gene = results$gene,
                      pval_adjusted = results$padj)
save(results, file = "output/lineage2_results.RData")
```

### Percentage of genes called significant as we change q-value threshold

```{r fig.width=5, fig.height=5, out.width='50%'}
propSig(objLP)
```

### Heatmap of significant genes

Scaled log expression of lineage 2. Heatmap of top 500 significant genes detected by three methods.

```{r, message=FALSE, fig.width=10, fig.height=7, out.width='100%'}
par(mfrow=c(1, 3))
layout(matrix(c(1,2,3), 1), widths = c(2, 2.7))
brk = sigGenesHeatmap(objLP, 2, TRUE, 500, "LineagePulse")
sigGenesHeatmapLoess(PvalP.adj, 2, TRUE, 500, "Loess Parametric", brk, TRUE)
# sigGenesHeatmapLoess(PvalNP.adj, 2, TRUE, 500, "Loess Nonparametric", brk, TRUE)
```

Log expression of lineage 2. Heatmap of top 500 significant genes detected by three methods.

```{r, message=FALSE, fig.width=10, fig.height=7, out.width='100%'}
par(mfrow=c(1, 3))
layout(matrix(c(1,2,3), 1), widths = c(2, 2.7))
brk = sigGenesHeatmap(objLP, 2, FALSE, 500, "LineagePulse")
sigGenesHeatmapLoess(PvalP.adj, 2, FALSE, 500, "Loess Parametric", brk, TRUE)
# sigGenesHeatmapLoess(PvalNP.adj, 2, FALSE, 500, "Loess Nonparametric", brk, TRUE)
```

### Visualization of 9 most significant genes

```{r fig.width=20, fig.height=12, out.width='100%'}
plotGenes(objLP, 9, "mostSig")
```

### Visualization of 9 least significant genes

```{r fig.width=20, fig.height=12, out.width='100%', warning=FALSE}
plotGenes(objLP, 9, "leastSig")
```

### Venn Diagram

```{r}
nGenes = 500
lp = objLP$dfResults
lp = lp[order(lp$padj), ]
lp = lp[1:nGenes, "gene"]
loessP = PvalP.adj[order(PvalP.adj$L2), ]
loessP = loessP[1:nGenes, "gene"]
loessNP = PvalNP.adj[order(PvalNP.adj$L2), ]
loessNP = loessNP[1:nGenes, "gene"]

  tmp <- venn.diagram(
    x = list(
      "LineagePulse" = lp, 
      "Loess" = loessP
    ),
    filename = paste0("Lineage2.png"),
        output = TRUE ,
        imagetype="png" ,
        compression = "lzw",
   col = "black",
    lty = "dotted",
    lwd = 2.5,
        height = 700 , 
        width = 700 , 
        resolution = 300,
        fill = c("cornflowerblue", "green"),    
    alpha = 0.40,
    cat.cex = 0.6,
   cex = 1.5,
    main.cex = 1.5,
    sub.cex = 1.25,
    cat.fontfamily = "serif"
  )
```

```{r}
nGenes = 500
lp = objLP$dfResults
lp = lp[lp$padj <= 0.01, "gene"]
loessP = PvalP.adj[PvalP.adj$L2 <= 0.01, "gene"]

  tmp <- venn.diagram(
    x = list(
      "LineagePulse" = lp, 
      "Loess" = loessP
    ),
    filename = paste0("Lineage2Sig.png"),
        output = TRUE ,
        imagetype="png" ,
        compression = "lzw",
   col = "black",
    lty = "dotted",
    lwd = 2.5,
        height = 700 , 
        width = 700 , 
        resolution = 300,
        fill = c("cornflowerblue", "green"),    
    alpha = 0.40,
    cat.cex = 0.6,
   cex = 1.5,
    main.cex = 1.5,
    sub.cex = 1.25,
    cat.fontfamily = "serif"
  )
```


# Observation

- Most of the genes (~70%) are called significant under a commonly used significant level 0.01.
- `LineagePulse` works well in detecting time-varying genes. However, it is very sensitive to outliers: several of the most significant genes are detected simply due to outliers (?) as shown in the heatmap. 
